---
html_document:
  df_print: paged
author: "Shawn P. Serbin"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
title: A basic PLSR example using leaf-level specrta and leaf mass per area (LMA) data from several CONUS NEON sites
---

```{r setup, include=FALSE}
knitr::opts_hooks$set(out.maxwidth <- function(options) {
  if (!knitr:::is_html_output()) return(options)
  options$out.extra <- sprintf('style="max-width: %s; margin: auto; border: none; display:block;"', options$out.maxwidth)
  options
})
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning = FALSE,
                      fig.align="center",
                      results = "asis",
                      out.maxwidth='100%')
outdir <- tempdir()
knitr::opts_knit$set(root.dir=outdir)
#library("xtable")
#head <- function(...) {
#  print(xtable(utils::head(...)), type="html")
#}
```

### Overview
his is an [R Markdown](http://rmarkdown.rstudio.com) Notebook to illustrate how to conduct a basic model fit.  This example shows you how to retrieve a dataset from the EcoSIS spectral database, choose the "optimal" number of plsr components, and fit a plsr model for leaf-mass area

When you click the **Knit** button in Rstudio a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

### Getting Started
### Installation
```{r, eval=FALSE }
list.of.packages <- c("readr","httr","pls","dplyr","reshape2")  # packages needed for script
# check for dependencies and install if needed
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```


##### Load libraries
```{r, echo=TRUE}
# load libraries needed for script
library(pls)
library(readr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(dplyr)
```

##### Prepare helpers
```{r, echo=TRUE}
# define function to grab PLSR model from GitHub
#devtools::source_gist("gist.github.com/christophergandrud/4466237")
source_GitHubData <-function(url, sep = ",", header = TRUE) {
  require(httr)
  request <- GET(url)
  stop_for_status(request)
  handle <- textConnection(content(request, as = 'text'))
  on.exit(close(handle))
  read.table(handle, sep = sep, header = header)
}

# not in
`%notin%` <- Negate(`%in%`)


# Script options
pls.options(plsralg = "oscorespls")
pls.options("plsralg")
pls.options()$parallel
# NULL
```

##### Setup temporary folder
```{r, echo=TRUE}
print(paste0("Output temporary directory: ",outdir))
setwd(outdir) # set working directory
getwd()  # check wd
```

### Grab data from EcoSIS
##### URL:  https://ecosis.org/package/fresh-leaf-spectra-to-estimate-lma-over-neon-domains-in-eastern-united-states
```{r, echo=TRUE}
print("**** Downloading Ecosis data ****")
ecosis_id <- "5617da17-c925-49fb-b395-45a51291bd2d"  # NEON dataset
ecosis_file <- sprintf(
  "https://ecosis.org/api/package/%s/export?metadata=true",
  ecosis_id
)
message("Downloading data...")
dat_raw <- read_csv(ecosis_file)
message("Download complete!")
head(dat_raw)
names(dat_raw)[1:40]
```

### Create PLSR dataset
```{r, echo=TRUE}
Start.wave <- 500
End.wave <- 2400
wv <- seq(Start.wave,End.wave,1)

spectra <- data.frame(dat_raw[,names(dat_raw) %in% wv])
names(spectra) <- c(paste0("Wave_",wv))
head(spectra)[1:6,1:10]
sample_info <- dat_raw[,names(dat_raw) %notin% seq(350,2500,1)]
head(sample_info)

sample_info2 <- sample_info %>%
  select(Domain,Functional_type,Sample_ID,USDA_Species_Code=`USDA Symbol`,LMA_gDW_m2=LMA)
head(sample_info2)

plsr_data <- data.frame(sample_info2,Spectra=I(as.matrix(spectra)))

inVar <- "LMA_gDW_m2"
```

### Plot the spectra
```{r, echo=TRUE}
cexaxis <- 1.5
cexlab <- 1.8
ylim <- 75

mean_spec <- colMeans(spectra[,which(names(spectra) %in% paste0("Wave_",wv))])
spectra_quantiles <- apply(spectra[,which(names(spectra) %in% paste0("Wave_",wv))],
                           2,quantile,na.rm=T,probs=c(0,0.025,0.05,0.5,0.95,0.975,1))

print("**** Plotting specrtal data ****")
par(mfrow=c(1,1), mar=c(4.5,5.7,0.3,0.4), oma=c(0.3,0.9,0.3,0.1)) # B, L, T, R
plot(wv,mean_spec,ylim=c(0,ylim),cex=0.00001, col="white",xlab="Wavelength (nm)",
     ylab="Reflectance (%)",cex.axis=cexaxis, cex.lab=cexlab)
polygon(c(wv ,rev(wv)),c(spectra_quantiles[5,]*100, rev(spectra_quantiles[3,]*100)),
        col="#99CC99",border=NA)
lines(wv,mean_spec*100,lwd=3, lty=1, col="black")
lines(wv,spectra_quantiles[1,]*100,lwd=1.85, lty=3, col="grey40")
lines(wv,spectra_quantiles[7,]*100,lwd=1.85, lty=3, col="grey40")
legend("topright",legend=c("Mean reflectance","Min/Max", "95% CI"),lty=c(1,3,1),
       lwd=c(3,3,15),col=c("black","grey40","#99CC99"),bty="n", cex=1.7)
box(lwd=2.2)
```

### Run Jackknife test to find number of components - simple example
```{r, echo=TRUE}
dims <- dim(plsr_data)
nComps <- 20
iterations <- 20
seg <- 5
prop <- 0.70
jk.out <- matrix(data=NA,nrow=iterations,ncol=nComps) 
pls.options(parallel = parallel::detectCores()-1) # Use mclapply
print("*** Running jacknife permutation test.  Please hang tight, this can take awhile ***")
start.time <- Sys.time()
for (i in 1:iterations) {
  rows <- sample(1:nrow(plsr_data),floor(prop*nrow(plsr_data)))
  sub.data <- plsr_data[rows,]
  plsr.out <- plsr(as.formula(paste(inVar,"~","Spectra")), scale=FALSE, center=TRUE, ncomp=nComps, 
                   validation="CV", segments = seg, segment.type="interleaved", trace=FALSE, data=sub.data)
  resPRESS <- as.vector(plsr.out$validation$PRESS)
  jk.out[i,seq(plsr.out$validation$ncomp)]=resPRESS
}
end.time <- Sys.time()
end.time - start.time
```

### PRESS plot
```{r, echo=TRUE}
pressDF <- as.data.frame(jk.out)
names(pressDF) <- as.character(seq(nComps))
pressDFres <- melt(pressDF)
bp <- ggplot(pressDFres, aes(x=variable, y=value)) + theme_bw() + 
  geom_boxplot(notch=TRUE) + labs(x="Number of Components", y="PRESS") +
  theme(axis.text=element_text(size=18), legend.position="none",
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))
bp
```

### Calculate optimal number of components
```{r, echo=TRUE}
# conduct t.test across components to identify first minimum - just one of the ways to do this
j <-2 
results <- as.vector(array(data="NA", dim=c(nComps-1,1)))
for (i in seq_along(1:nComps-1)) {
  comp1 <- i; comp2 <- j
  ttest <- t.test(pressDFres$value[which(pressDFres$variable==comp1)],
                    pressDFres$value[which(pressDFres$variable==comp2)])
  #print(i)
  results[i] <- round(unlist(ttest$p.value),8)
  j <- j+1
  if (j > nComps) {
    break
  }
}
results <- data.frame(seq(2,nComps,1),results)
names(results) <- c("Component", "P.value")
results

# *** based on t.test - optimal components are 16 ***
# NOTE: Becuase the jacknife test above depends on random selection
# the optimal components may change slightly between differe runs of this script
# This is expected given different permutatoins and uncertainty in the data 
```

### Final PLSR model fit
```{r, echo=TRUE}
# Simple final model validated with cross-validation.  Segmented cross-validation used
# given the very large sample size.  For models with fewer observations (e.g. <100) 
# LOO or leave-one-out cross validation is recommended

#nComps <- 14
first <- min(which(as.numeric(as.character(results$P.value)) > 0.05))
nComps <- results$Component[first]
print(paste0("*** Optimal number of components based on t.test: ", nComps))

segs <- 30
pls.options(parallel = NULL)
plsr.out <- plsr(as.formula(paste(inVar,"~","Spectra")),scale=FALSE,ncomp=nComps,validation="CV",
                 segments=segs, segment.type="interleaved",trace=TRUE,data=plsr_data)
fit <- plsr.out$fitted.values[,1,nComps]

plot_data <- data.frame(plsr_data[, which(names(plsr_data) %notin% "Spectra")], Fitted=fit)
plot_data <- plot_data %>%
  mutate(Residuals = Fitted-LMA_gDW_m2)
head(plot_data)
```

### Basic scatter plot of results
```{r, echo=TRUE}
scatter_plot <- ggplot(plot_data, aes(x=Fitted, y=LMA_gDW_m2)) + 
  theme_bw() + geom_point() + geom_abline(intercept = 0, slope = 1, color="dark grey", 
            linetype="dashed", size=1.5) + xlim(0, 275) + ylim(0, 275) +
  labs(x=expression(paste("Predicted LMA (",g~m^{-2},")")), 
       y=expression(paste("Observed LMA (",g~m^{-2},")"))) +
  annotate("text", x=250, y=70, label = paste0("R^2 == ", round(pls::R2(plsr.out)[[1]][nComps],2)), parse=T) + 
  annotate("text", x=250, y=40, label = paste0("RMSE == ", round(pls::RMSEP(plsr.out)[[1]][nComps],2)), parse=T) +
  theme(axis.text=element_text(size=18), legend.position="none",
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))
scatter_plot

resid_histogram <- ggplot(plot_data, aes(x=Residuals)) +
  geom_histogram(binwidth=.5, alpha=.5, position="identity") + 
  geom_vline(xintercept = 0, color="black", 
              linetype="dashed", size=1) + theme_bw() + 
  theme(axis.text=element_text(size=18), legend.position="none",
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))
resid_histogram
```


### Scatter plot by Functional_type
```{r, echo=TRUE}
scatter_plot <- ggplot(plot_data, aes(x=Fitted, y=LMA_gDW_m2)) + 
  theme_bw() + geom_point(aes(fill=Functional_type),alpha=0.6,colour="black", pch=21, size=4) + 
  geom_abline(intercept = 0, slope = 1, color="dark grey", 
              linetype="dashed", size=1.5) + xlim(0, 275) + ylim(0, 275) +
  labs(x=expression(paste("Predicted LMA (",g~m^{-2},")")), 
       y=expression(paste("Observed LMA (",g~m^{-2},")"))) +
  annotate("text", x=250, y=70, label = paste0("R^2 == ", round(pls::R2(plsr.out)[[1]][nComps],2)), parse=T) + 
  annotate("text", x=250, y=40, label = paste0("RMSE == ", round(pls::RMSEP(plsr.out)[[1]][nComps],2)), parse=T) +
  theme(axis.text=element_text(size=18), legend.position="bottom",legend.title=element_text(size=16),
        legend.text=element_text(size=14),
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))
scatter_plot

resid_histogram <- ggplot(plot_data, aes(x=Residuals, fill=Functional_type)) +
  geom_histogram(binwidth=.5, alpha=.5, position="identity") + 
  geom_vline(xintercept = 0, color="black", alpha=0.6,
             linetype="dashed", size=1) + theme_bw() + 
  theme(axis.text=element_text(size=18), legend.position="bottom",legend.title=element_text(size=16),
        legend.text=element_text(size=14),
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))
resid_histogram
```

### Scatter plot by Domain
```{r, echo=TRUE}
scatter_plot <- ggplot(plot_data, aes(x=Fitted, y=LMA_gDW_m2)) + 
  theme_bw() + geom_point(aes(fill=Domain),alpha=0.6,colour="black", pch=21, size=4) + 
  geom_abline(intercept = 0, slope = 1, color="dark grey", 
              linetype="dashed", size=1.5) + xlim(0, 275) + ylim(0, 275) +
  labs(x=expression(paste("Predicted LMA (",g~m^{-2},")")), 
       y=expression(paste("Observed LMA (",g~m^{-2},")"))) +
  annotate("text", x=250, y=70, label = paste0("R^2 == ", round(pls::R2(plsr.out)[[1]][nComps],2)), parse=T) + 
  annotate("text", x=250, y=40, label = paste0("RMSE == ", round(pls::RMSEP(plsr.out)[[1]][nComps],2)), parse=T) +
  theme(axis.text=element_text(size=18), legend.position="bottom",legend.title=element_text(size=16),
        legend.text=element_text(size=14),
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))
scatter_plot

resid_histogram <- ggplot(plot_data, aes(x=Residuals, fill=Domain)) +
  geom_histogram(binwidth=.5, alpha=.5, position="identity") + 
  geom_vline(xintercept = 0, color="black", alpha=0.6,
             linetype="dashed", size=1) + theme_bw() + 
  theme(axis.text=element_text(size=18), legend.position="bottom",legend.title=element_text(size=16),
        legend.text=element_text(size=14),
        axis.title=element_text(size=20, face="bold"), 
        axis.text.x = element_text(angle = 0,vjust = 0.5),
        panel.border = element_rect(linetype = "solid", fill = NA, size=1.5))
resid_histogram
```

###### eof